---
title: "Democracy Reports"
date: 2022-07-07T10:00:00-06:00
publishDate: 2022-07-07T10:00:00-06:00
draft: false

# Authors. Comma separated list, e.g. `["Bob Smith", "David Jones"]`.
authors: [admin]

# Abstract and optional shortened version.
abstract: ""
summary: "How do democracy ratings vary around the world? I use FreedomHouse data to find out."

# Tags (optional).
#   Set `tags: []` for no tags, or use the form `tags: ["A Tag", "Another Tag"]` for one or more tags.
tags: ["R Projects"]

# Links (optional).
url_pdf: ""
url_slides: ""
url_video: ""
url_code: ""

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(kableExtra)


fh_wb <- read_csv(here("content", "research", "democracy-reports", "files", "fh_wb.csv"))

all_countries <- unique(fh_wb$country)

allfiles <- list.files("files")
goodfiles <- grep(list.files("files"), pattern = "_files|.csv", invert = TRUE)


dt_url <- vector(mode = "character", length = length(goodfiles))
count = 1
for (i in goodfiles) {
  dt_url[count] <- allfiles[i]
  count = count+1
}

dt_url <- paste("files/", dt_url, sep = "")

all_countries <- all_countries %>%
                    sort()

```

## Democracy in the World
How does the health of democracy vary from country to country within a given continent? Using [FreedomHouse](https://freedomhouse.org/report/freedom-world) data from 1973-2019, I generated country profiles for all instances where data was available. The profiles also incorporate economic indicators from the [World Bank](https://data.worldbank.org/indicator).  

I am currently working to expand this analysis to include measurements of inequality, and to explore the relationship between income and democracy on a global scale.  

These profiles are fully reproducible. [Contact me](https://davidicrabtree.com#contact) for the R scripts.  


```{r, echo=FALSE}


all_countries %>% 
  cell_spec(row.names(.), "html", link = dt_url) %>%
  kable("html", escape = FALSE, col.names =  "Country Profile") %>%
  kable_styling(bootstrap_options = c("hover", "striped", "condensed")) %>%
  scroll_box(width = "100%", height = "800px")

```



