---
title: "Tidy Data"
subtitle: "What It Is and How To Do It"  
author: 
  - "David I. Crabtree"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

knitr::opts_chunk$set(
  cache = TRUE,
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8, fig.asp = 0.618, fig.retina = 2, dpi = 150,
  fig.showtext = TRUE,
  fig.align = "center",
  out.width = "80%"
  
)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#E22C6C",
  secondary_color = "#3E33CC",
  inverse_header_color = "#FFFFFF"
)
```

## What is Tidy Data?

#### Consider these two dataframes. Which one looks easier to work with? 

--

.pull-left[
##### Frame 1
```{r, echo=FALSE}
library(here)
library(tidyverse)
tidydata <- read_csv(here("static", "slides", "tidying-data", "fh_wb.csv"))

head(select(tidydata, 1, 4, 5))


```
]

--

.pull-right[
##### Frame 2
```{r, echo=FALSE}
library(here)
library(readxl)
untidydata <- read_excel(here("static", "slides", "tidying-data", "FreedomHouse.xlsx"), 
                 sheet = 2, # read in data from the second sheet
                 na = "-", # recode "-" to missing
                 skip = 1) 

head(select(untidydata, 1, 2, 5))
```
]

--

<br> 

They might seem similar at first glance, but the frame on the left will be significantly easier to work with. In the following slides I'll explain why. 

---

.pull-left[
## What is Tidy Data?

##### There are three rules that all Tidy Data must follow:

1. **Each variable gets its own column.**  
1. **Each observation gets its own row.**  
1. **Each value gets its own cell.** 

Every tidy dataset will follow these rules.  
]

--

.pull-right[

### Consider an example

```{r}
head(starwars[1:4])
```
]



---

## What is Tidy Data: Which of these tables is tidy? 

.pull-left[

```{r}
table1[1:4,]
table2[1:4,]
```

]  

.pull-right[
```{r}
table3[1:4,]
```

#### Example continues on next slide

]

---

## What is Tidy Data: Which of these datasets is tidy? 



.pull-left[
```{r}
table4a[1:3,] # cases
```
]

.pull-right[
```{r}
table4b[1:3,]  # population
```
]

---

## What is Tidy Data?

#### Table 1 is the only table that is tidy.

--

```{r}
table1[1:4,]
```

--

### Note why this data is tidy
(1) Every variable has its own column.  
(2) Every observation has its own row.  
(3) Every value has its own cell.  

---

class: inverse, middle, center

# Tidying Data: How to Do It 

---

## Tidying Data

#### So what do we do if data has broken one (or more!) of the rules of tidy data? 

--

#### Well, we bring it back into compliance!

--

In order to tidy - or wrangle - the data, we will use the following functions from the `dplyr` package. 

* `pivot_longer()`
* `pivot_wider()` 
* `separate()`
* `unite()`

---

## Tidying Data: `pivot_longer()`

#### When data has observations that should be columns, use `pivot_longer()`.  

--

Usually, this means that the dataset has column names that are not _names_ of variables but _values_ of variables; this violates the first rule of tidy data.  
<br>
Consider an example of untidy data from the previous slides

--

```{r, echo=FALSE}
table4a[1:3,]
```

--

1999 and 2000 are not variable names; they're values of an unlisted variable - **year**. Thus, they should be in their own column. The `pivot_longer()` function achieves this. To do so, it requires three main arguments: which columns are to be moved (`cols`), what should be the new column that these ex-columns are stored in (`names_to`), and what should be the new column that the previously stored values go to (`values_to`)?

#### Try to solve this yourself before proceeding.

---

## Tidying Data: `pivot_longer()`

--

.pull-left[
```{r}
table4a
```
]

--

.pull-right[
```{r}
pivot_longer(data = table4a,
             cols = c(`1999`, `2000`),
             names_to = "year",
             values_to = "cases")
```

Note that the resulting dataframe is now tidy: **Each variable in its own column**, each observation in its own row, each value in its own cell.

The **unit of analysis** is the case count of a given country in a given year. 

]

---

## Tidying Data: `pivot_wider()`

#### When data has columns that should be observations, use `pivot_wider()`.

--

Usually, this will be because one observation is spread across multiple rows instead of just one; this violates the second rule of tidy data.  

Consider another example we've looked at previously: 

--

```{r, echo=FALSE}
table2[1:4,]
```

--

Note that in this table, the **unit of analysis** is no longer "the case count of a given country in a given year," like it was in our tidy dataframe. Now, the unit of analysis is "a given type of count" in a given country in a given year. Wow - that's a loy of "given"s! And it is precisely why Afghanistan is listed four times here!  

There is a row for Afghanistan's cases in 1999, a row for its population in 1999, a row for its cases in 2000, a row for its population in 2000, and so on. To tidy it, we need to move rows to be columns.

---

## Tidying Data: `pivot_wider()`

--

.pull-left[
```{r}
table2
```
]

--

.pull-right[
```{r}
pivot_wider(data = table2,
            names_from = "type",
            values_from = "count")
```

Note that the resulting dataframe is now tidy: Each variable in its own column, **each observation in its own row**, each value in its own cell.

The **unit of analysis** is the case count of a given country in a given year. 

]

---

## Tidying Data: `separate()`

#### When data has cells with more than one value, use `separate()`.

--

Usually, this will involve the creation of a new column to store the new values from the over-filled cell. 

<br>
Consider `table3` from earlier: 

--

```{r, echo=FALSE}
table3[1:4,]
```

Note that the cells in the rate column actually contain two values: the case count first, then the population count. This violates the third rule of tidy data. To tidy it, each value should have its own cell.  

Can you guess what the new columns will be called? 

---

## Tidying Data: `separate()`

--

.pull-left[
```{r}
table3
```
]

--

.pull-right[
```{r}
separate(data = table3,
         col = "rate",
         into = c("cases", "population"),
         sep = "/")
```

Note that the resulting dataframe is now tidy: Each variable in its own column, each observation in its own row, **each value in its own cell**.

The **unit of analysis** is the case count of a given country in a given year. 

]

---

## Tidying Data: `unite()`

#### When data has cells with *less* than one value, use `unite()`.

--

This is rare, but it occurs when a meaningful value is spread across multiple columns. This violates the third rule of tidy data. 

<br>
Consider `table5`: 

--

```{r, echo=FALSE}
table5[1:4,]
```

--

Note that a meaningful value of year is obscurred by the columns `century` and `year`. Instead of having Afghanistan in 1999, we have Afghanistan in the 99th year of the 19th century (... which, come to think of it, would be 1899!). `Unite()` fixes this by combining the two columns 

Can you guess what the new column will be called? 

---

## Tidying Data: `unite()`

--

.pull-left[
```{r untidy table5}
table5
```
]

--

.pull-right[
```{r}
unite(data = table5,
      col = "year",
      century, year)
```

Note that the new year column is not formatted correctly. This is because it is now separated with a "_". We can fix this with the `sep = ` argument. 

]

---

## Tidying Data: `unite()`

.pull-left[
```{r untidy table5}
```
]

.pull-right[
```{r}
unite(data = table5,
      col = "year",
      century, year,
      sep = "")
```

Note that the resulting dataframe is now tidy: Each variable in its own column, each observation in its own row, **each value in its own cell**.

The **unit of analysis** is the case count of a given country in a given year. 
]

---


# Conclusion

Tidy data follows three core principles: every variable gets its own column, every observation gets its own row, every value gets its own cell.  

To clean our data into tidy data, we can use the following functions from the `dplyr` package:
- `pivot_longer()` when observations are columns
- `pivot_wider()` when columns are observations
- `separate()` when a cell has more than 1 value
- `unite()` when a cell has less than 1 meaningful value

<br>

#### References

Wickam, Hadley and Garrett Grolemund, [*R for Data Science.*](https://r4ds.had.co.nz/index.html)

---

class: center, middle, inverse

# Questions?

## Email me! 

### crabtreed@uchicago.edu
